---
title: "mini-project"
author: "James Lee"
date: "11/12/2020"
output: pdf_document
---

# 1. Exploratory data analysis

## Preparing the data



```{r}
# Save your input data file into your Project directory
fna.data <- "WisconsinCancer.csv"

# Complete the following code to input the data and store as wisc.df
wisc.df <- read.csv(fna.data, row.names=1)

head(wisc.df)
```

```{r}
# We can use -1 here to remove the first column
wisc.data <- wisc.df[,-1]

```


```{r}
# Create diagnosis vector for later 
diagnosis <- wisc.df$diagnosis

```



## Exploratory data analysis

### Q1. How many observations are in this dataset?
```{r}
nrow(wisc.df)
```
569 observation 


### Q2. How many of the observations have a malignant diagnosis?
```{r}
nrow(wisc.df[which(wisc.df$diagnosis == "M"),])
```


### Q3. How many variables/features in the data are suffixed with _mean?
```{r}
head(wisc.df)

```

10 variables are suffixed with _mean



# 2. Principal Component Analysis
## Performing PCA


```{r}
# Check column means and standard deviations
colMeans(wisc.data)

apply(wisc.data,2,sd)

```

```{r}
# Perform PCA on wisc.data by completing the following code
wisc.pr <- prcomp(wisc.data, scale = TRUE)
summary(wisc.pr)

```


###  Q4. From your results, what proportion of the original variance is captured by the first principal components (PC1)?
44.27%

### Q5. How many principal components (PCs) are required to describe at least 70% of the original variance in the data?
the first 3 PCs

### Q6. How many principal components (PCs) are required to describe at least 90% of the original variance in the data?
the first 7 PCs


## Interpreting PCA results

```{r}
biplot(wisc.pr)
```


### Q7. What stands out to you about this plot? Is it easy or difficult to understand? Why?
THe plot is too compact and all the words are on top of each other so it is too difficult to understand. This is mainly because we have too many variables and 30 different PCs. 


```{r}
# Scatter plot observations by components 1 and 2
plot(wisc.pr$x[,1], wisc.pr$x[,2], col = diagnosis , 
     xlab = "PC1", ylab = "PC2")

```



### Q8. Generate a similar plot for principal components 1 and 3. What do you notice about these plots?
```{r}
plot(wisc.pr$x[,1], wisc.pr$x[,3], col = diagnosis , 
     xlab = "PC1", ylab = "PC3")

```


Both plots have two groups of clusters in the same area. But since PC2 has more variance in the data, the first plot has more distinct clusters than the second plot. 





```{r}
# Create a data.frame for ggplot
df <- as.data.frame(wisc.pr$x)
df$diagnosis <- diagnosis

# Load the ggplot2 package
library(ggplot2)

# Make a scatter plot colored by diagnosis
ggplot(df) + 
  aes(PC1, PC2, col=diagnosis) + 
  geom_point()

```


## Variance explained

```{r}
# Calculate variance of each component
pr.var <- wisc.pr$sdev^2
head(pr.var)

```

```{r}

# Variance explained by each principal component: pve
pve <- pr.var / sum(pr.var)

# Plot variance explained for each principal component
plot(pve, xlab = "Principal Component", 
     ylab = "Proportion of Variance Explained", 
     ylim = c(0, 1), type = "o")
```


```{r}
# Alternative scree plot of the same data, note data driven y-axis
barplot(pve, ylab = "Precent of Variance Explained",
     names.arg=paste0("PC",1:length(pve)), las=2, axes = FALSE)
axis(2, at=pve, labels=round(pve,2)*100 )

```

```{r}
## ggplot based graph
#install.packages("factoextra")
library(factoextra)
fviz_eig(wisc.pr, addlabels = TRUE)

```


## Communicating PCA results

### Q9. For the first principal component, what is the component of the loading vector (i.e. wisc.pr$rotation[,1]) for the feature concave.points_mean?
```{r}
wisc.pr$rotation[,1]
```

-0.26085376

### Q10. What is the minimum number of principal components required to explain 80% of the variance of the data?
the first 5 PCs



# 3. OPTIONAL: Hierarchical clustering

```{r}
# Scale the wisc.data data using the "scale()" function
data.scaled <- scale(wisc.data)
data.dist <- dist(data.scaled)
wisc.hclust <- hclust(data.dist, method = "complete")

```

## Results of hierarchical clustering

### Q11. Using the plot() and abline() functions, what is the height at which the clustering model has 4 clusters?

```{r}
plot(wisc.hclust)
abline(wisc.hclust, col="red", lty=2)
```


## Selecting number of clusters

```{r}
wisc.hclust.clusters <- cutree(wisc.hclust, k = 4)

table(wisc.hclust.clusters, diagnosis)

```



# 5. Combining methods
## Clustering on PCA results


```{r}

wisc.pr.hclust <- hclust(dist(wisc.pr$x[, 1:7]), method = "ward.D2")
grps <- cutree(wisc.pr.hclust, k=2)
table(grps)

```


```{r}
table(grps, diagnosis)

```

```{r}
plot(wisc.pr$x[,1:2], col=grps)

```


```{r}
plot(wisc.pr$x[,1:2], col=diagnosis)

```




```{r}
g <- as.factor(grps)
levels(g)

```

```{r}
g <- relevel(g,2)
levels(g)

```


```{r}
plot(wisc.pr$x[,1:2], col=g)
```

```{r}
library(rgl)
plot3d(wisc.pr$x[,1:3], xlab="PC 1", ylab="PC 2", zlab="PC 3", cex=1.5, size=1, type="s", col=grps)

```




# 7. Prediction

## Clustering on PCA results
```{r}
url <- "https://tinyurl.com/new-samples-CSV"
new <- read.csv(url)
npc <- predict(wisc.pr, newdata=new)
npc

```

```{r}
plot(wisc.pr$x[,1:2], col=g)
points(npc[,1], npc[,2], col="blue", pch=16, cex=3)
text(npc[,1], npc[,2], c(1,2), col="white")

```


## Q17. Which of these new patients should we prioritize for follow up based on your results?

Patient 2 (red dots) should be prioritized since that patient has the malignant tumor

